\subsection{Referencias}

NREL – Eagle Supercomputer
National Renewable Energy Laboratory. (2019, January 8). NREL’s New Supercomputer, 
Eagle, Takes Flight. National Renewable Energy Laboratory. 
 
 https://www.nrel.gov/manufacturing/news/program/2019/nrel-new-supercomputer-eagle-takes-flight#:~:text=Now%20in%20production%20use%2C%20the,accelerate%20our%20energy%20system%20transformation www2.nrel.gov

\noindent\rule{\textwidth}{0.4pt}

Top500 – June 2025 List
TOP500. (2025, junio). TOP500 Supercomputer Sites – June 2025. TOP500. 
 
https://top500.org/lists/top500/2025/06/ top500.org

\noindent\rule{\textwidth}{0.4pt}

Glenn Klockwood – Eagle (Microsoft)
Klockwood, G. (s. f.). Eagle (Microsoft). In The HPC Gardener. 
 
https://glennklockwood.com/garden/systems/Eagle glennklockwood.com

\noindent\rule{\textwidth}{0.4pt}

Glenn Klockwood – Azure ND H100 v5 Node
Klockwood, G. (s. f.). Azure ND H100 v5. In The HPC Gardener. 
 
https://glennklockwood.com/garden/nodes/Azure-ND-H100-v5 glennklockwood.com

\noindent\rule{\textwidth}{0.4pt}

Microsoft Learn – Dasv6 Series
Microsoft Corporation. (2024, noviembre 21). Dasv6 sizes series. Microsoft Learn. 
 
https://learn.microsoft.com/en-us/azure/virtual-machines/sizes/general-purpose/dasv6-series?tabs=sizebasic Microsoft Learn

\noindent\rule{\textwidth}{0.4pt}

Microsoft Learn – Msv2 Medium Memory Series
Microsoft Corporation. (2025, junio 6). Msv2 Medium Memory size series. Microsoft Learn. 
 
https://learn.microsoft.com/en-us/azure/virtual-machines/sizes/memory-optimized/msv2-mm-series?tabs=sizebasic Microsoft Learn

\noindent\rule{\textwidth}{0.4pt}

Microsoft Learn – Msv3 High Memory Series
Microsoft Corporation. (2025, junio 6). Msv3 High Memory size series. Microsoft Learn. 

https://learn.microsoft.com/en-us/azure/virtual-machines/sizes/memory-optimized/msv3-hm-series?tabs=sizebasic Microsoft Learn
 